Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-06-22 analysis/paper/paper.Rmd
Found and saved SSPD as ADMUR pd and exponential, logistic and uniform MLE models

The procedure outlined above represent the standard approach when modelling population dynamics in archaeology. This can be helpful for determining whether the population behaviour has followed the trajectories of standard theoretical models, and is a useful first step here as few assumptions concerning the behaviour and overall development of number of shoreline dated sites over time could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs themselves. As a result, and by improving on @goldstein2016, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics by use of RSPDs, published with the accompanying R package  *ADMUR*. This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure of model comparison, and finally evaluating the best model using a goodness-of-fit test in line with the Monte Carlo approach outlined above. The benefit of the CPL models is that these define discrete linear periods of development between hinge points that mark points in time in which the trajectory of the population proxy changes direction, thus allowing for the identification of key events with a conservative description of population dynamics between these. The approach thus also provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model. Following the evaluation of standard models as outlined above, the data were therefore subsequently subjected to the approach presented by @timpson2021.

@@

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in a version-controlled online repository at X. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017]. Analysis of the ^14^C-dates were done using the R package *ADMUR* [@timpson2021]. The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with repurposed code from both *ADMUR* and *rcarbon* [@crema2021] for Monte Carlo simulation and model comparison.

@@

To account for investigatory bias that can result from variable sampling intensity between sites, the radiocarbon dates were initially binned using the *binPrep()* function from *rcarbon*, where dates were aggregated on a site by site basis if they fell within 200 uncalibrated ^14^C years of each other. When calibrated, these are summed and divided by the number of dates in the bin before being included in the final RSPD. Furthermore, the ^14^C-dates were not normalised. This follows from a peculiarity in the calibration procedure which can lead the probability mass of a date to sum to more than one [@weninger2015]. As a consequence of this, normalising the dates before summing has been shown to create spurious spikes in the RSPD [e.g. @crema2021]. All calibration was done using the IntCal20 calibration curve [@reimer2020].

The RSPD was first subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE search, using the differential evolution optimization algorithm *DEoptimR*. The Monte Carlo simulations were then performed using the *modelTest()* function from *rcarbon*. For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10000 times for each null model. The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the critical envelope is then used to calculate a global p-value, indicating whether or not the null can be rejected.

@@

Following from the commutative nature of the shoreline dates dealt with here, the Monte Carlo simulation with the SSPD is here based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any individual year is determined by the null model of choice. This is equivalent to the sampling method *calsample* from *rcarbon* [@crema2021]. Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2 km interval. The calendar date is then 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is then drawn with uniform probability between the lower and upper limit of the displacement curve at that calendar year, using intervals of 5 cm. As shoreline dating is here done with a gamma distribution with the same parameters across all sites, there is no equivalent of the error term for ^14^C-dates that determine the shape of the Gaussian distribution associated with the ^14^C age. Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was then shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples. Having dated the number of samples equalling the number of shoreline dated sites, these were then summed, and the entire process repeated a total of 10000 times. The 97.5% highest and 2.5% lowest summed probability at each interval of 5 years across all simulations was then retrieved to create the 95% critical envelope, and the global p-value found in the same manner as for the RSPDs. This was done using re-purposed code from *rcarbon*.

@@

```{r spduniform, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline dated sites (n = 934) compared to a uniform null model. B) Same as A, but with a exponential model. C) Summed probability of calibrated radiocarbon dates (n =, bins = 678) compared to a uniform null model. D) Same as C, but with a exponential null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
Comparing the the directions and deviations from the three standard models, Figure appears to indicate some process of overall general increase or uniform distribution in the frequency of radiocarbon dates over time, contrasted to a general decrease in shoreline dated sites over time. However, with global p-values all below X,  all three models can be rejected as explaining the data they have been fit to, both for the SSPD and RSPD. Based on common demographic developments and previous suggestions, there is possibly some theoretical justification for expecting an exponential or logistic model to explain the shape of the RSPD, thus giving some further credence to interpreting deviations from this overall development as potentially meaningful signals of demographic developments. However, as introduced above, several authors have warned against over-interpreting such deviation. Furthermore, there are even less foundations on which to hypothesise what processes have dictated the development of site frequency over time, thus offering even further challenges to the interpretation of the SSPD and it's relationship with the null-models.

@@

While it appears that the SSPD and RSPD are differentially impacted by population numbers, and, we suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation. Some lines of inquiry could be to also draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @riede2012; @solheim2020], and quantitative measures for mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @roalkvam2022]. Furthermore, environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of importance to unpack their relationship [e.g. @ordonez2022; @bergsvik2022; @jorgensen2020].

@@

Furthermore, seeing as the method for shoreline dating implemented here was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties associated with its application [see @roalkvam2023]. While we still believe the approach represents a considerable improvement over previous approaches, the results should

>>>>

The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs. This can be helpful for determining whether the population behaviour has followed the trajectories of standard theoretical models, and is a useful first step here also for the treatment of shoreline dated sites, as few assumptions concerning the verall development of this proxy could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs themselves. As a result, and by improving on @goldstein2016, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics by use of RSPDs, published with the accompanying R package  *ADMUR*. This involves fitting continuous piece-wise linear (CPL) models to the data, in addition to the standard models, subjecting these to a procedure of model comparison, and finally evaluating the best model using a goodness-of-fit test, in line with the Monte Carlo approach outlined above. The CPL models are defined by discrete linear periods of development between hinge points that mark points in time in which the trajectory of the population proxy changes direction. This therefore allows for the identification of key events and a conservative description of population dynamics between these. The approach thus also provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model. Following the evaluation of standard models as outlined above, the data were therefore subsequently subjected to the approach presented by @timpson2021.

@@

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in a version-controlled online repository at X. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017]. Analysis of the ^14^C-dates were done using the R package *ADMUR* [@timpson2021]. The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* [@crema2021] for Monte Carlo simulation and model comparison.

@@

To account for investigatory bias that can result from variable sampling intensity between sites, the summing of radiocarbon dates with *ADMUR* starts with binning dates from the same sites that fall within 200 uncalibrated ^14^C years of each other when measured from the mean ^14^C age. All dates are then calibrated using a resoltuion of 5 years and then normalised to sum to unity. The RSPDs for each site-phase are then summed to achieve the final SPD, which is in turn normalised. All calibrations were performed using the IntCal20 calibration curve [@reimer2020].

The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE search, using the differential evolution optimization algorithm *DEoptimR*. The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*. For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10000 times for each null model. The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the critical envelope is then used to calculate a global p-value, indicating whether or not the null can be rejected.

@@

The Monte Carlo simulations with the SSPD is here based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any 5 year interval is determined by the null model of choice. This is equivalent to the sampling method *calsample* from *rcarbon* [@crema2021]. Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2 km interval. The calendar date is then 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is then drawn with uniform probability between the lower and upper limit of the displacement curve at that calendar year, using intervals of 5 cm. As shoreline dating is here done with a gamma distribution with the same parameters across all sites, there is no equivalent of the error term for ^14^C-dates that determine the shape of the Gaussian distribution associated with the ^14^C age. Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was then shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples. Having dated the number of samples equalling the number of shoreline dated sites, these were then summed, and the entire process repeated a total of 10000 times. The 97.5% highest and 2.5% lowest summed probability at each interval of 5 years across all simulations was then retrieved to create the 95% critical envelope, and the global p-value found in the same manner as for the RSPDs. This was done using re-purposed code from *rcarbon*.

@@

```{r spduniform, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline dated sites (n = 934) compared to a uniform null model. B) Same as A, but with a exponential model. C) Summed probability of calibrated radiocarbon dates (n =, bins = 678) compared to a uniform null model. D) Same as C, but with a exponential null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
For the RSPD, the logistic model achieved a p-value of, and could therefore not be rejected. Based on the available data, this therefore represents the best explanation of the data. We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case.

Comparing the the directions and deviations from the three standard models, Figure appears to indicate that some process of overall general increase or uniform distribution in the frequency of radiocarbon dates over time, contrasted to a general decrease in shoreline dated sites over time. However, with global p-values all below X,  all three models can be rejected as explaining the data they have been fit to, both for the SSPD and RSPD. Based on common demographic developments and previous suggestions, there is possibly some theoretical justification for expecting an exponential or logistic model to explain the shape of the RSPD, thus giving some further credence to interpreting deviations from this overall development as potentially meaningful signals of demographic developments. However, as introduced above, several authors have warned against over-interpreting such deviation. Furthermore, there are even less foundations on which to hypothesise what processes have dictated the development of site frequency over time, thus offering even further challenges to the interpretation of the SSPD and it's relationship with the null-models.

@@

While it appears that the SSPD and RSPD are differentially impacted by population numbers, and, we suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation. Some lines of inquiry could be to also draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @riede2012; @solheim2020], and quantitative measures for mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @roalkvam2022]. Furthermore, environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack their relationship [e.g. @ordonez2022; @bergsvik2022; @jorgensen2020].

Finally, seeing as the method for shoreline dating implemented here was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties associated with its application [see @roalkvam2023]. While we still believe the approach represents an improvement over previous approaches, the results should therefore be treated with care. This also extends to the results of the RSPD, due to the fairly limited ^14^C sample. With these limitations in mind, the approach outlined by @timpson2021 as it offers a more conservative treatment of the data, reducing the amount of over-interpretation that might simply be the result of uncertainties to do with data or methods.

@@
